### 2021-02-22 HW-AdvancedQFT2020/advqft1/advqft1.tex
Update note on the quantum boost generator

	\item Promote $
		\var{X}\mapsto \epsilon(x)\var{X}
		= \epsilon(x)\,\pqty{
			a^\mu + \lambda\id{^\mu_\nu} X^\nu
		}
	$, with $\epsilon(x)$ some localized bump function; using \eqref{eq:lorentz_variation} and considering \textit{on-shell} variation, we have:
	\begin{equation}
	\begin{aligned}
		0 = \var{S}
		&= - \int \dd[2]{x} \pqty{
			\pd^\alpha X_\mu\,
				a^\mu\,
				\pdd{\alpha}\epsilon
			+ X^\nu\,\pd^\alpha X^\mu\,
				\lambda_{\mu\nu}\,
				\pdd{\alpha}\epsilon
		} \\
		&= - \int \dd[2]{x} \pqty{
			\pd^\alpha X_\mu\,
				a^\mu\,
			+ X_{[\nu}\,\pd^\alpha X_{\mu]}\,
				\lambda^{[\mu\nu]}\,
		}\,\pdd{\alpha}\epsilon
	\end{aligned}
	\end{equation}
	It is evident (after partial integration) that the following currents are conserved; they are the Noether currents associated with $a^\mu$ and $\lambda^{[\mu\nu]}$:
	\begin{equation}
		j^\alpha_\mu
		= -\pd^\alpha X_\mu,\quad
		j^\alpha_{\mu\nu}
		= -X_{[\nu}\,\pd^\alpha X_{\mu]}
		= \frac{1}{2}\,\pqty{
			X_\mu\,\pd^\alpha X_\nu
			- X_\nu\,\pd^\alpha X_\mu
		}
	\end{equation}

@@

	We've obtained an interesting result: the expression for the boost generator $M^{i0}$ contains the Hamiltonian density $\mcal{H}$, weighted by the radial distance $x^i$. This is natural since a boost does indeed contains time evolution for states away from the origin. It's an important result utilized by the so-called \textit{Rindler decomposition}; in fact, $M^{i0}$ becomes the Hamiltonian for an accelerated observer in the Rindler patch\footnote{
		See the lecture notes of Tom Hartman: \http{hartmanhep.net/topics2015/gravity-lectures.pdf} or Daniel Harlow \arxiv{1409.1231}. 
	}.

	For $M^{ij}$, we have:
	\begin{equation}
		M^{ij}
		= - \int \dd[3]{x}
			\dot{\phi}\,\pqty{
				x^i \pd^j
				- x^j \pd^i
			}\,\phi
	\end{equation}
	This is interpreted as the angular momentum of the field $\phi$. Suppose $\phi$ is a wave packet localized around $\vb{x}$ with momentum $\approx \vb{p}$, then we have the classical angular momentum up to some factor:
	\begin{equation}
		M^{ij}
		\sim \pqty{
				x^i p^j
				- x^j p^i
			} \int \dd[3]{x}
				E \phi^2,
	\quad
		E = \sqrt{\vb{p}^2 + m^2}
	\end{equation}

	The $
		\int \dd[3]{x} E \phi^2
	$ factor in the above expression is an $\order{1}$ normalization constant for a particle-like wave packet; to see this, note that $\phi\in\mbb{R}$ has a phase factor $
		\phi
		\sim a\,e^{+ik\cdot x} + a^\dagger e^{-ik\cdot x}
		\sim \cos\,(k\cdot x)
	$,
	\begin{gather}
	\small
		E = \int \dd[3]{x} \mcal{H}
		= \int \dd[3]{x} \pqty{
				\frac{1}{2} \dot{\phi}^2
				+ \frac{1}{2} (\grad{\phi})^2
				+ \frac{1}{2} m^2 \phi^2
				+ \cdots
			}
		\sim \pqty{E^2 + \vb{p^2} + m^2 + \cdots}
			\int \dd[3]{x} \frac{1}{2}\,\phi^2,
	\end{gather}
	\vspace{-.7\baselineskip}
	\begin{equation}
		E\int \dd[3]{x} \phi^2 \sim 1,
	\end{equation}
	Indeed, we have: $
		M^{ij}
		\sim \pqty{
				x^i p^j
				- x^j p^i
			}
	$.

	Canonical quantization:
	\begin{gather}
		[\phi(\vb{x}),\varpi(\vb{y})]
		= [\phi(\vb{x}),\dot{\phi}(\vb{y})]
		= i\delta(\vb{x} - \vb{y})
	\end{gather}
	Other equal-time commutators between $\phi,\varpi$ all just vanish. Operator products are then regularized by normal ordering: $M\mapsto \normorder{M}$, which can be explicitly implemented by normal ordering of the oscilator modes:
	\begin{equation}
		\phi(x)
		= \int \frac{\dd[3]{k}}{(2\pi)^3}
			\frac{1}{\sqrt{2E_k}}
			\pqty{
				a_{\vb{k}} e^{ik\cdot x}
				+ a^\dagger_{\vb{k}'} e^{-ik\cdot x}
			},
	\quad
		[a_{\vb{k}}, a^\dagger_{\vb{k}'}]
		= (2\pi)^3 \delta(\vb{k} - \vb{k}')
	\end{equation}

@@

	We note that this result is almost the classical $
		x^0 P^i - x^i P^0
	$, but here $x^i P^0$ is replaced with the integral over energy density $\mcal{H}$. The result can be nicely re-written with the stress tensor $T^{\mu\nu}$; just run the Noether's procedure with $\var{x^\mu} = \epsilon^\mu$, then we shall obtain:
	\begin{equation}
	\begin{gathered}
		j'^\mu
		= \epsilon_\nu T^{\mu\nu},
	\quad
		T^{\mu\nu}
		= \pd^\mu\phi\,\pd^\nu\phi
			+ \eta^{\mu\nu} \mcal{L},
	\\
		Q' = \epsilon_\mu P^\mu,
	\quad
		P^\mu
		= \int \dd[3]{x} T^{\mu 0}
		= \int \dd[3]{x} \pqty{
			\pd^0\phi\,\pd^\mu\phi
			+ \eta^{0\mu} \mcal{L}
		}
	\\
		M^{\mu\nu}
		= \int \dd[3]{x}
			2x^{[\mu} T^{\nu]0}
	\end{gathered}
	\end{equation}
	The quantization $M\mapsto \normorder{M}$ is thus reduced to the quantization of $T_{0\nu}$.

>>>>

	\item Promote $
		\var{X}\mapsto \epsilon(x)\var{X}
		= \epsilon(x)\,\pqty{
			a^\mu + \lambda\id{^\mu_\nu} X^\nu
		}
	$, with $\epsilon(x)$ some localized bump function; using \eqref{eq:lorentz_variation} and considering \textit{on-shell} variation, we have:
	\begin{equation}
	\begin{aligned}
		0 = \var{S}
		&= - \int \dd[2]{x} \pqty{
			\pd^\alpha X_\mu\,
				a^\mu\,
				\pdd{\alpha}\epsilon
			+ X^\nu\,\pd^\alpha X^\mu\,
				\lambda_{\mu\nu}\,
				\pdd{\alpha}\epsilon
		} \\
		&= - \int \dd[2]{x} \pqty{
			\pd^\alpha X_\mu\,
				a^\mu\,
			+ X_{[\nu}\,\pd^\alpha X_{\mu]}\,
				\lambda^{[\mu\nu]}\,
		}\,\pdd{\alpha}\epsilon
	\end{aligned}
	\end{equation}
	It is then evident (after partial integration) that the following currents are conserved; they are the Noether currents associated with $a^\mu$ and $\lambda^{[\mu\nu]}$:
	\begin{equation}
		j^\alpha_\mu
		= -\pd^\alpha X_\mu,\quad
		j^\alpha_{\mu\nu}
		= -X_{[\nu}\,\pd^\alpha X_{\mu]}
		= \frac{1}{2}\,\pqty{
			X_\mu\,\pd^\alpha X_\nu
			- X_\nu\,\pd^\alpha X_\mu
		}
	\end{equation}

@@

	We've obtained an interesting result: the expression for the boost generator $M^{i0}$ contains the Hamiltonian density $\mcal{H}$, weighted by the radial distance $x^i$. This is natural since a boost does indeed contains time evolution for excitations away from the origin. It's an important result utilized by the so-called \textit{Rindler decomposition}; in fact, $M^{i0}$ becomes the Hamiltonian for an accelerated observer in the Rindler patch\footnote{
		See the lecture notes of Tom Hartman: \http{hartmanhep.net/topics2015/gravity-lectures.pdf} or Daniel Harlow \arxiv{1409.1231}. 
	}.

	For $M^{ij}$, we have:
	\begin{equation}
		M^{ij}
		= - \int \dd[3]{x}
			\dot{\phi}\,\pqty{
				x^i \pd^j
				- x^j \pd^i
			}\,\phi
	\end{equation}
	This is interpreted as the angular momentum of the field $\phi$. Suppose $\phi$ is a wave packet localized around $\vb{x}$ with momentum $\approx \vb{k}$, then we have the classical angular momentum up to some factor:
	\begin{equation}
		M^{ij}
		\sim \pqty{
				x^i k^j
				- x^j k^i
			} \int \dd[3]{x}
				E \phi^2,
	\quad
		E = \sqrt{\vb{k}^2 + m^2}
	\end{equation}

	The $
		\int \dd[3]{x} E \phi^2
	$ factor in the above expression is an $\order{1}$ normalization constant for a particle-like wave packet; to see this, note that $\phi\in\mbb{R}$ has a phase factor $
		\phi
		\sim a\,e^{+ik\cdot x} + a^\dagger e^{-ik\cdot x}
		\sim \cos\,(k\cdot x)
	$,
	\begin{gather}
	\small
		E = \int \dd[3]{x} \mcal{H}
		= \int \dd[3]{x} \pqty{
				\frac{1}{2} \dot{\phi}^2
				+ \frac{1}{2} (\grad{\phi})^2
				+ \frac{1}{2} m^2 \phi^2
				+ \cdots
			}
		\sim \pqty{E^2 + \vb{k^2} + m^2 + \cdots}
			\int \dd[3]{x} \frac{1}{2}\,\phi^2,
	\end{gather}
	\vspace{-.7\baselineskip}
	\begin{equation}
		E\int \dd[3]{x} \phi^2 \sim 1,
	\end{equation}
	Indeed, we have: $
		M^{ij}
		\sim \pqty{
				x^i k^j
				- x^j k^i
			}
	$.

	Canonical quantization:
	\begin{gather}
		[\phi(\vb{x}),\varpi(\vb{y})]
		= [\phi(\vb{x}),\dot{\phi}(\vb{y})]
		= i\delta(\vb{x} - \vb{y})
	\end{gather}
	Other equal-time commutators between $\phi,\varpi$ all just vanish. Operator products are then regularized by normal ordering: $M\mapsto \normorder{M}$, which can be explicitly implemented by normal ordering of the oscilator modes:
	\begin{equation}
		\phi(x)
		= \int \frac{\dd[3]{k}}{(2\pi)^3}
			\frac{1}{\sqrt{2E_k}}
			\pqty{
				a_{k} e^{ik\cdot x}
				+ a^\dagger_{k} e^{-ik\cdot x}
			},
	\quad
		[a_{k}, a^\dagger_{k'}]
		= (2\pi)^3 \delta(\vb{k} - \vb{k}')
	\end{equation}
	The $k$ dependence in $a_k, E_k$ is, in fact, only a $\vb{k}$ dependence; we've dropped the boldface in the subscripts simply for convenience.

@@

	We note that this result is almost the classical $
		x^0 P^i - x^i P^0
	$, but here $x^i P^0$ is replaced with the integral over energy density $\mcal{H}$. The result can be nicely re-written with the stress tensor $T^{\mu\nu}$; just run the Noether's procedure with $\var{x^\mu} = \epsilon^\mu$, then we shall obtain:
	\begin{equation}
	\begin{gathered}
		j'^\mu
		= \epsilon_\nu T^{\mu\nu},
	\quad
		T^{\mu\nu}
		= \pd^\mu\phi\,\pd^\nu\phi
			+ \eta^{\mu\nu} \mcal{L},
	\\
		Q' = \epsilon_\mu P^\mu,
	\quad
		P^\mu
		= \int \dd[3]{x} T^{\mu 0}
		= \int \dd[3]{x} \pqty{
			\pd^0\phi\,\pd^\mu\phi
			+ \eta^{0\mu} \mcal{L}
		}
	\\
		M^{\mu\nu}
		= \int \dd[3]{x}
			2x^{[\mu} T^{\nu]0}
	\end{gathered}
	\end{equation}
	The quantization $M\mapsto \normorder{M}$ is thus reduced to the quantization of $T^{\nu 0}$, weighted by a $x^\mu$ factor.

	First let's look at $T^{00} = \mcal{H}$; note that:
	\begin{equation}
		\int \dd[3]{x}
			x^i e^{i\vb{k}\cdot\vb{x}}
		= (2\pi)^3 \pqty{
				-i\,\pdv{k_i}
			} \delta(\vb{k})
	\end{equation}
	One can then check explicitly with mode expansion that, up to normal ordering, we have\footnote{
		See a similar result in \url{https://physics.stackexchange.com/q/27906}. Moreover, the charge can be computed at arbitrary time slice $t$, but the $t$-dependence ($\sim e^{\pm iEt}$) drops out in the final result, due to the on-shell condition $E_k^2 = \vb{k}^2 + m^2$ and symmetries, e.g.\ $\int \dd[3]{k} k^i = 0$. Note that $\pdv{E_k}{k_i} = \frac{k^i}{E_k}$. 
	}:
	\begin{gather}
		H = \int \dd[3]{x} \mcal{H}
		= \int \frac{\dd[3]{k}}{(2\pi)^3}\,
			E_k\, a^\dagger_k a_k,
	\\
		\mcal{O}^i = \int \dd[3]{x} x^i \mcal{H}
		= \int \frac{\dd[3]{k}}{(2\pi)^3}\,
			E_k\, a^\dagger_k
				\pqty{+i\,\pdv{k_i}}
			a_k,
	\end{gather}

%\pagebreak[3]
	At first glance, derivative of $a_k = a_{\vb{k}}$ with respect to $k^i$ seems puzzling; however, note that $\pqty\big{+i\,\pdv{k_i}}$ is precisely the $x^i$ operator in ``momentum-space'', and one can make sense of it by considering a generic $n$-particle state:
	\begin{equation}
		\ket{\Psi}
		= \int \dd[3]{k_1}\cdots \dd[3]{k_n}
			\Psi(\vb{k}_1,\cdots,\vb{k}_n)
			\,a^\dagger_{k_1}
			\cdots a^\dagger_{k_n} \ket{0}
	\end{equation}
	Where $\Psi(\vb{k}_1,\cdots,\vb{k}_n)$ is the $n$-particle wave function; using $
		a_k a^\dagger_{k'}
		= a^\dagger_{k'} a_k
			+ (2\pi)^3\delta(\vb{k} - \vb{k'})
	$ recursively, we get in the end that:
	\begin{equation}
		\mcal{O}_i \ket{\Psi}
		= \int \dd[3]{k_1}\cdots \dd[3]{k_n}
			\Bqty{
				\sum_{m = 1}^n
				E_{k_m}
				\pqty{+i\,\pdv{k^i_m}}\,
				\Psi(\vb{k}_1,\cdots,\vb{k}_n)
			}
			\,a^\dagger_{k_1}
			\cdots a^\dagger_{k_n} \ket{0}
	\end{equation}

	We see that indeed $\mcal{O}_i$ acts as $
		E_k \pqty{+i\,\pdv{k^i}}
	$ on the momentum-space $n$-particle wave function, consistent with the result of ordinary quantum mechanics; ...
